
---
title: "Prediction of Gender Unemployment Rates for 2023"
author: "Madeleine Ashby"
date: "9/13/2022"
output: html_document
fig_width: 6 
fig_height: 4 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Import necessary libraries / packages
library(readxl)
library(dplyr)
library(ggplot2)
library(tstools)
library(zoo)
```

# Exploratory Data Analysis

## Data Collection
The data we are using was obtained from the [Bureau of Labor Statistics] (https://data.bls.gov/cgi-bin/surveymost?ln) which provides labor force statistics from the current United States population survey.  We selected two categories of data -- unemployment rate for women 20 years and older, and unemployment rate for men 20 years and older -- from the year 1950 to present.

The resulting dataset is a multi-series table that gives the unemployment rate (as a number representing a percent) for both men and women each month from January 1950 to August 2022.

```{r}
unemployment_data <- read_excel("C:/Users/Maddie/Documents/DS 4002/Project 1/Unemployment.xlsx")
```

## Data Cleaning
```{r}
# Transpose the data to be more easily readable
unemployment_data <- t(unemployment_data)

# Make sure it's actually a dataframe to be able to use functions on it in the future
unemployment_data <- as.data.frame(unemployment_data)

# Change column names
unemployment_data <-
  unemployment_data %>%
  rename(
    Male = V1,
    Female = V2
  )

# Remove first row with "Series ID" by subsetting
unemployment_data <- unemployment_data[-1, ]

View(unemployment_data)
```

## Preliminary Analysis

### Univariate Analysis
```{r}
male_time_series <- ts(unemployment_data$Male, start = c(1950, 1), end = c(2022, 8), frequency=12)
#print(male_time_series)
plot(male_time_series,
     main = "Male Unemployment Rate in the United States from 1950 to 2022",
     xlab = "Time",
     ylab = "Unemployment Rate"
     )

female_time_series <- ts(unemployment_data$Female, start = c(1950, 1), end = c(2022, 8), frequency=12)
#print(female_time_series)
plot(female_time_series,
     main = "Female Unemployment Rate in the United States from 1950 to 2022",
     xlab = "Time",
     ylab = "Unemployment Rate"
     )

```

### Bivariate Analysis
```{r}
tsplot(Male=male_time_series, Female=female_time_series,
       plot_title = "Unemployment Rate in the United States from 1950 to 2022: Male vs Female",
       auto_legend = TRUE
       )
```


# Analysis

## ARIMA
```{r}
unemployment_data$Female <- as.double(unemployment_data$Female)
unemployment_data$Male <- as.double(unemployment_data$Male)

# Female ARIMA
female_ARIMA <- auto.arima(unemployment_data$Female)
female_ARIMA

#residuals
plot.ts(female_ARIMA$residuals)

female_prediction <- forecast(female_ARIMA, level=c(95), h=12)
plot(female_prediction)

#BoxL-jung Test
Box.test(female_prediction$resid, lag=5, type="Ljung-Box")

# Male ARIMA
male_ARIMA <- auto.arima(unemployment_data$Male)
male_ARIMA

#residuals
plot.ts(male_ARIMA$residuals)

male_prediction <- forecast(male_ARIMA, level=c(95), h=12)
plot(male_prediction)

#BoxL-jung Test
Box.test(male_prediction$resid, lag=5, type="Ljung-Box")

```


